---
title: "16S internal control log"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        toc: true
        fig_width: 10
        fig_height: 6
params:
    emu: '' # For docker
    #emu: "data/emu-combined-species-counts.tsv" # For testing locally
    github: ''
    MOCK_SPP: !r c('Staphylococcus saprophyticus', 'Cutibacterium avidum', 'Streptococcus equi', 'Imtechella halotolerans', 'Allobacillus halotolerans', 'Truepera radiovictrix')
    STDEV_NO: 20
    MAX_LABELS: 10
    MAX_SAMPLES: 35
---

<style>
   tbody tr:nth-child(odd){
    background-color: #F7FBFF;
  }
  th, td { padding: 5px; }
</style>

```{r dependencies, include = FALSE}
library(vegan)
library(knitr)
source("/usr/local/src/rscripts/ordiellipse_custom_scale.R") # For Docker
#source("rscripts/ordiellipse_custom_scale.R") # For testing locally
environment(ordiellipse_custom_scale) = asNamespace('vegan')
```

### Samples

```{r data, echo=FALSE}
STDEV_NO = params$STDEV_NO
MAX_LABELS = params$MAX_LABELS
MAX_SAMPLES = params$MAX_SAMPLES
MOCK_SPP = params$MOCK_SPP

count_table = read.csv(params$emu, header=TRUE, sep="\t")
row.names(count_table)=count_table$species
count_table = count_table[,8:ncol(count_table)]
count_table[is.na(count_table)] = 0
colnames(count_table) = sub("X", "", colnames(count_table)) # remove X

# Sorting by name 
count_table = count_table[, order(names(count_table))]

# Factors
factors = as.data.frame(names(count_table))
row.names(factors) = factors$`names(count_table)`
factors$`names(count_table)` = NULL
factors$sample_no = as.character(1:nrow(factors))
factors$no_reads = colSums(count_table)
factors$no_taxa = specnumber(count_table, MARGIN=2)

# Add sample type when >n samples
if (length(row.names(factors)) >= STDEV_NO) {
  factors$sample = as.factor(c(rep("stdev", STDEV_NO),rep("test",(length(row.names(factors))-STDEV_NO))))
  test_samples = names(count_table) %in% row.names(factors)[factors$sample=="test"] # Get test samples
  if (length(which(test_samples == TRUE)) > MAX_LABELS) {
    test_subset = names(count_table) %in% row.names(factors)[c((nrow(factors)-MAX_LABELS+1):nrow(factors))]
    use_subset = TRUE
  } else {
    use_subset = FALSE
  }
}

# Calculate proportions from counts
prop_table = prop.table(as.matrix(count_table), 2)

```

Standard deviation is calculated from the first `r STDEV_NO` samples.\
Total number of samples: `r length(row.names(factors))`\
Number of test samples: `r length(which(factors$sample == "test"))`

```{r table, echo=FALSE}
# Print sample list and type
kable(factors, format="html", align="l")
```

### PCA

```{r pca, echo=FALSE}
samples.rda = rda(t(prop_table) ~ 1)
coldict = c("stdev"="white","test"="red")

if (length(row.names(factors)) >= STDEV_NO) {
  # 2STDEV stdev and test samples
  plot(samples.rda, display="sites", type="n", main="2 standard deviations")
  with(factors, points(samples.rda, display="sites", pch=21, bg=as.character(coldict[sample])))
  with(factors, ordiellipse_custom_scale(samples.rda, sample, show.groups="stdev", kind="sd", scale.custom=2, draw="polygon", col="skyblue", border="blue"))
  with(factors, legend("topright", legend=levels(sample), bty="n", pch=21, col="black", pt.bg=coldict))
  if (length(which(test_samples) == TRUE) > 0 & length(which(test_samples) == TRUE) < 3) {
    with(factors, text(samples.rda, display="sites", select=test_samples, cex=0.75))
  }
    
  # 3STDEV stdev and test samples
  plot(samples.rda, display="sites", type="n", main="3 standard deviations")
  with(factors, points(samples.rda, display="sites", pch=21, bg=as.character(coldict[sample])))
  with(factors, ordiellipse_custom_scale(samples.rda, sample, show.groups="stdev", kind="sd", scale.custom=3, draw="polygon", col="skyblue", border="blue"))
  with(factors, legend("topright", legend=levels(sample), bty="n", pch=21, col="black", pt.bg=coldict))
  if (length(which(test_samples) == TRUE) > 0 & length(which(test_samples) == TRUE) < 3) {
    with(factors, text(samples.rda, display="sites", select=test_samples, cex=0.75))
  }

  # 2 STDEV - only show test samples with stdev ellipses
  if (length(which(test_samples) == TRUE) > 2) {
    if (use_subset == TRUE) {
      ordipointlabel(samples.rda, select=test_subset, display="sites", cex=0.75, pch=21, bg="red")
    } else {
      ordipointlabel(samples.rda, select=test_samples, display="sites", cex=0.75, pch=21, bg="red")
    }
    with(factors, ordiellipse_custom_scale(samples.rda, sample, show.groups="stdev", kind="sd", scale.custom=2, draw="polygon", col="skyblue", border="blue"))
    with(factors, legend("topright", legend="test", bty="n", pch=21, col="black", pt.bg="red"))
    title(main=paste0("2 standard deviations - max. ", MAX_LABELS, " of the latest test samples"))
  }

  # 3STDEV - only show test samples with stdev ellipses
    if (length(which(test_samples) == TRUE) > 2) {
    if (use_subset == TRUE) {
      ordipointlabel(samples.rda, select=test_subset, display="sites", cex=0.75, pch=21, bg="red")
    } else {
      ordipointlabel(samples.rda, select=test_samples, display="sites", cex=0.75, pch=21, bg="red")
    }
    with(factors, ordiellipse_custom_scale(samples.rda, sample, show.groups="stdev", kind="sd", scale.custom=3, draw="polygon", col="skyblue", border="blue"))
    with(factors, legend("topright", legend="test", bty="n", pch=21, col="black", pt.bg="red"))
    title(main=paste0("3 standard deviations - max. ", MAX_LABELS, " of the latest test samples"))
  }

} else if (length(row.names(factors)) > 2) {
    if (length(row.names(factors)) >= 8) {
      plot(samples.rda, display="sites", type="points", pch=21, main=paste0("PCA - less than ", STDEV_NO, " samples"))
      ordipointlabel(samples.rda, display="sites", cex=0.75, pch=21)
    }
    else {
      plot(samples.rda, display="sites", type="text", main=paste0("PCA - less than ", STDEV_NO, " samples"))
    }

} else {
  cat("Too few samples for PCA")
}

```

### Relative abundance of mock species
A maximum of `r MAX_SAMPLES` of the latest test sampels are shown.

```{r mock species, echo=FALSE}
mock_table = prop_table[row.names(prop_table) %in% MOCK_SPP,]
mock_table = mock_table[MOCK_SPP,]*100 # transform to percent

# Restrict no of samples shown in mock taxa plots to for example 35
if (length(row.names(factors)) > (STDEV_NO + MAX_SAMPLES)) {
  mock_table = mock_table[,c(1:STDEV_NO,(ncol(mock_table)-MAX_SAMPLES+1):ncol(mock_table))]
}

for (i in 1:length(dimnames(mock_table)[[1]])) {
  par(mar=c(18,6,2,0), oma=c(0, 0, 0, 0))
  layout(matrix(c(1,1,1,1,1,2), nrow=1, ncol=6, byrow=TRUE))
  if (length(row.names(factors)) >= STDEV_NO) {
    sd_samples = sd(mock_table[i,which(factors$sample == "stdev")])
    mean_samples = mean(mock_table[i,which(factors$sample == "stdev")])
    
    plot(mock_table[i,], ylab=paste0("% ", row.names(mock_table)[i]), xlab="", xaxt="n", ylim=c(mean_samples-(3*sd_samples),mean_samples+(3*sd_samples)), pch=21, bg=c("white", "red")[factors$sample], cex=1.5, cex.lab=1.5)
    axis(1, labels=dimnames(mock_table)[[2]], at=1:ncol(mock_table), las=2, cex=1.5, cex.axis=1)

    abline(h=mean_samples+(3*sd_samples), lty=3)
    abline(h=mean_samples+(2*sd_samples), lty=2)
    abline(h=mean_samples)
    abline(h=mean_samples-(2*sd_samples), lty=2)
    abline(h=mean_samples-(3*sd_samples), lty=3)
    
    plot.new()
    par(mar=c(1,1,1,1), oma=c(0, 0, 0, 0))
    legend('topright', legend=c("3 sd", "2 sd"), lty=c(3,2), cex=1.5)
    legend('bottomright', legend=levels(factors$sample), pch=21, pt.bg=c("white","red"), cex=1.5)
  }
  # If not enough samples to calculate stdev: plot without SD lines
  else {
    plot(mock_table[i,], ylab=paste0("% ", row.names(mock_table)[i]), xlab="", xaxt="n", pch=21, bg=c("white"), cex=1.5, cex.lab=1.5)
    axis(1, labels=dimnames(mock_table)[[2]], at=1:ncol(mock_table), las=2, cex=1.5, cex.axis=1)
    
    plot.new()
    par(mar=c(1,1,1,1), oma=c(0, 0, 0, 0))
    legend('bottomright', legend=c("stdev"), pch=21, pt.bg=c("white"), cex=1.5)
  }
}
```

### Software

This report was generated with:\
`r params$github`\
R `r getRversion()`\
R markdown `r packageVersion('rmarkdown')`\
vegan `r packageVersion('vegan')`
